% !TEX root =main.tex
\section{Derivative of Regular Expressions with Fuzzy Similarity Relation}\label{sec:fuzzy-derivatives}
In this paper, we generalize this notion using fuzzy similarity. Suppose $\cR$ is
a fuzzy similarity relation on $\Sigma$, and let $\cut$ be a cut value
($0 < \cut \le 1$). A first possible generalization only compares the symbol
being consumed with the first symbol of a word in the language. This is too
narrow for a similarity-based setting: after the first symbol is consumed, the
remaining suffix should also be compared up to the same cut similarity. We
therefore define the fuzzy derivative by matching the whole word up to
$\cut$-similarity: the head symbol is matched against $a$, and the returned
suffix is allowed to be any suffix similar to the actual remaining suffix.

\begin{definition}\label{fuzzyderivative}
Let $\mathcal{L}\subseteq\Sigma^*$ and $a\in\Sigma$. The \emph{$\cut$-fuzzy
derivative} of $\mathcal{L}$ with respect to $a$ is
$$
\partial_a^\cut(\mathcal{L})
=
\{w \in \Sigma^* \mid
  \exists b \in \Sigma,\ \exists v\in\Sigma^*
  \text{ such that }
  \cR(a,b) \ge \cut,\ \mathcal{R}^w(w,v)\ge\cut,\ \text{and } bv \in \mathcal{L}
\}.
$$
Alternatively:
$$
\partial_a^\cut(\mathcal{L})
=
\{w \in \Sigma^* \mid
  \exists b \in \Sigma,\ \exists v\in\Sigma^*
  \text{ such that }
  (a,b)\in\cR_\cut,\ w\sim_\cut^{\Sigma^*}v,\ \text{and } bv\in\mathcal{L}
\}.
$$
\end{definition}

\begin{proposition}\label{prop:cut-equivalence}
Let $\cR$ be a fuzzy similarity relation on $\Sigma$ and let $\cut$ be a cut value
with $0 < \cut \le 1$. Define
\[
  a \sim_\cut^\Sigma b \quad\Longleftrightarrow\quad \cR(a,b)\ge \cut.
\]
Then $\sim_\cut^\Sigma$ is an equivalence relation on $\Sigma$. It induces an
equivalence relation on words by
\[
  a_1\cdots a_n \sim_\cut^{\Sigma^*} b_1\cdots b_m
  \quad\Longleftrightarrow\quad
  n=m \text{ and } a_i \sim_\cut^\Sigma b_i \text{ for every } i.
\]
\end{proposition}

\begin{definition}\label{def:quotient-language-fuzzy}
Let $\Sigma^\cut := \Sigma/{\sim_\cut^\Sigma}$ and let
$q_\cut:\Sigma\to\Sigma^\cut$ be the quotient map, $q_\cut(a)=[a]_\cut$.
Extend $q_\cut$ to words by
\[
  q_\cut^*(a_1\cdots a_n) = [a_1]_\cut \cdots [a_n]_\cut.
\]
For a language $\mathcal L \subseteq \Sigma^*$, define its quotient image by
\[
  \mathcal L^\cut := q_\cut^*(\mathcal L)
  = \{q_\cut^*(w)\mid w\in\mathcal L\}
  \subseteq (\Sigma^\cut)^*.
\]
\end{definition}

\noindent
The map $q_\cut$ sends each symbol to its $\cut$-equivalence class. The map
$q_\cut^*$ does exactly the same thing for a whole word, letter by letter.

\smallskip
\noindent
\textbf{Example.} Suppose $\Sigma=\{a,b,c\}$ and, for the fixed cut $\cut$, we
have $a\sim_\cut^\Sigma b$ but $c$ is not equivalent to them. Then
\[
  q_\cut(a)=[a]_\cut,\qquad q_\cut(b)=[a]_\cut,\qquad q_\cut(c)=[c]_\cut.
\]
For the word $w=bac$ we obtain
\[
  q_\cut^*(bac)=[a]_\cut [a]_\cut [c]_\cut.
\]
So $q_\cut^*$ forgets which representative from a class was used and remembers
only the sequence of classes.

\smallskip
\noindent
Likewise, $\mathcal L^\cut$ is formed by replacing each word of $\mathcal L$ by
its class-word. For example, if
\[
  \mathcal L=\{aa,ba,c\},
\]
then
\[
  \mathcal L^\cut
  =
  \{[a]_\cut[a]_\cut,\ [a]_\cut[a]_\cut,\ [c]_\cut\}
  =
  \{[a]_\cut[a]_\cut,\ [c]_\cut\}.
\]
Thus different words of $\mathcal L$ can become the same word in the quotient
language $\mathcal L^\cut$.

\noindent
The next theorem explains how the fuzzy derivative becomes an ordinary derivative
after passing to the quotient alphabet: once letters are identified up to
$\cut$-similarity, differentiating with respect to $a$ is the same as
differentiating with respect to the class $[a]_\cut$.

{\color{green!50!black}
\begin{theorem}\label{the:quotient-fuzzy-derivative}
Let $\cR$ be a fuzzy similarity relation on $\Sigma$ and let $\cut$ be a cut value
with $0 < \cut \le 1$. For every language $\mathcal L \subseteq \Sigma^*$ and every
$a\in\Sigma$,
\[
  \partial_a^\cut(\mathcal L)
  \;=\;
  \{w\in\Sigma^* \mid q_\cut^*(w)\in
       \partial_{[a]_\cut}\bigl(\mathcal L^\cut\bigr)\}.
\]
Consequently,
\[
  q_\cut^*\bigl(\partial_a^\cut(\mathcal L)\bigr)
  =
  \partial_{[a]_\cut}\bigl(\mathcal L^\cut\bigr).
\]
\end{theorem}

\begin{proof}[Proof sketch]
We prove the two inclusions.

Let $w\in\partial_a^\cut(\mathcal L)$. By Definition~\ref{fuzzyderivative},
there are $b\in\Sigma$ and $v\in\Sigma^*$ such that $\cR(a,b)\ge\cut$,
$w\sim_\cut^{\Sigma^*}v$, and $bv\in\mathcal L$. Hence
$[b]_\cut=[a]_\cut$ and $q_\cut^*(w)=q_\cut^*(v)$, so
\[
  [a]_\cut q_\cut^*(w)
  =
  [b]_\cut q_\cut^*(v)
  =
  q_\cut^*(bv)
  \in \mathcal L^\cut.
\]
Thus $q_\cut^*(w)\in\partial_{[a]_\cut}(\mathcal L^\cut)$.

Conversely, suppose $q_\cut^*(w)\in\partial_{[a]_\cut}(\mathcal L^\cut)$. Then
$[a]_\cut q_\cut^*(w)\in\mathcal L^\cut$, so there is a word $bv\in\mathcal L$
such that
\[
  q_\cut^*(bv)=[a]_\cut q_\cut^*(w).
\]
It follows that $[b]_\cut=[a]_\cut$, hence $\cR(a,b)\ge\cut$, and also
$q_\cut^*(v)=q_\cut^*(w)$, hence $w\sim_\cut^{\Sigma^*}v$. Therefore
$w\in\partial_a^\cut(\mathcal L)$.

The displayed consequence follows because
$\partial_{[a]_\cut}(\mathcal L^\cut)$ consists only of quotient words, and each
such word has at least one representative in $\Sigma^*$.
\end{proof}
}

\medskip
\noindent\textbf{Relation with the head-only definition.}
The earlier head-only version was
\[
  \partial_{a,\mathrm{head}}^\cut(\mathcal L)
  =
  \{w\in\Sigma^* \mid \exists b\in\Sigma,\ (a,b)\in\cR_\cut
      \text{ and } bw\in\mathcal L\}.
\]
The new definition is exactly the $\cut$-saturation of this language:
\[
  \partial_a^\cut(\mathcal L)
  =
  \{w\in\Sigma^* \mid
       \exists v\in\partial_{a,\mathrm{head}}^\cut(\mathcal L)
       \text{ with } w\sim_\cut^{\Sigma^*}v\}.
\]
Equivalently, it is the inverse image of the ordinary quotient derivative:
\[
  \partial_a^\cut(\mathcal L)
  =
  (q_\cut^*)^{-1}
  \bigl(\partial_{[a]_\cut}(\mathcal L^\cut)\bigr).
\]

\begin{remark}[Counterexample]
Let $\Sigma=\{a,b\}$ and let $\cut=0.8$. Define a fuzzy similarity relation by
\[
  \cR(a,a)=\cR(b,b)=1,
  \qquad
  \cR(a,b)=\cR(b,a)=0.9.
\]
Then $a\sim_\cut^\Sigma b$, so $\Sigma^\cut$ has a single equivalence class,
namely $[a]_\cut=[b]_\cut$.

Take the language
\[
  \mathcal L=\{aa\}.
\]
For the symbol $b$, the fuzzy derivative is
\[
  \partial_b^\cut(\mathcal L)=\{a,b\},
\]
because $b$ is $\cut$-similar to the first $a$ in $aa$, and the returned suffix
may be any one-letter word $\cut$-similar to the actual suffix $a$.

On the other hand,
\[
  \mathcal L^\cut=\{[a]_\cut [a]_\cut\},
  \qquad
  \partial_{[b]_\cut}(\mathcal L^\cut)=\{[a]_\cut\}.
\]
This means that, at the quotient level, the only suffix we see is the
one-letter class-word $[a]_\cut$. Translating this back to ordinary words
collects all one-letter representatives of this class:
\[
  \{w\in\Sigma^* \mid q_\cut^*(w) \in \partial_{[b]_\cut}(\mathcal L^\cut)\}
  =
  \{a,b\},
\]
because both one-letter words $a$ and $b$ belong to the same class
$[a]_\cut=[b]_\cut$.

\smallskip
\noindent
More explicitly,
\[
  q_\cut^*(a)=[a]_\cut,
  \qquad
  q_\cut^*(b)=[a]_\cut,
\]
so both $a$ and $b$ are counted by the set
\[
  \{w\in\Sigma^* \mid q_\cut^*(w) \in \partial_{[b]_\cut}(\mathcal L^\cut)\}.
\]
This is now the intended result. Under the old head-only definition the result
would have been only $\{a\}$; the word $b$ was missing because the tail was not
matched fuzzily.
\end{remark}

\noindent
Just as in the classical setting, the notion of a fuzzy derivative with respect to a single symbol can be extended to handle derivatives with respect to an entire word. The idea remains the same: process the word one symbol at a time from left to right, taking the fuzzy derivative with respect to the first symbol and then applying the fuzzy derivative again to the result for the remaining suffix. This recursive approach naturally generalizes the single-symbol fuzzy derivative to arbitrary words, as formalized below.


\begin{definition}
    The fuzzy derivative of a language $\mathcal{L} \subseteq \Sigma^*$ with respect to a word $u=a\cdot v \in \Sigma^*$ where $a \in \Sigma$ and $v \in \Sigma^*$ is defined to be:
    $$
    \partial^\cut_u(\mathcal{L}) = \partial^\cut_v(\partial^\cut_a(\mathcal{L}))
    $$
\end{definition}


{\color{red} A more interesting theory arises if we take
$$
    \partial^{\mu\otimes\mu'}_u(\mathcal{L}) = \partial^{\mu'}_v(\partial^\cut_a(\mathcal{L}))
    $$

}
\noindent
An important property of the classical derivative is that it preserves regularity. This ensures that repeated differentiation never produces a language outside the regular class, making the operation suitable for automata-based  algorithms. The same closure property holds in the fuzzy setting: even when derivatives are taken with respect to a cut-based similarity relation, the resulting language remains regular. The following theorem states this formally.


\begin{theorem}\label{the:language-fuzzy-derivative-regular}
    If $\mathcal{L} \subseteq \Sigma^*$ is regular, then 
    $\partial^\cut_u(\mathcal{L})$ is also regular for any word 
    $u \in \Sigma^*$ and for any cut value $0 < \cut \le 1$.
\end{theorem}

\begin{proof}
The argument is similar to the proof of Theorem~\ref{the:language-derivative-regular}.
Let $\mathcal{L} \subseteq \Sigma^*$ be regular and fix $0 < \cut \le 1$.

For a single symbol $a \in \Sigma$ we have, by Definition of $\partial_a^\cut$,
\[
  \partial_a^\cut(\mathcal{L})
  =
  \{ w \in \Sigma^* \mid
    \exists b\in\Sigma,\ \exists v\in\Sigma^*
    \text{ with } \cR(a,b)\ge\cut,\ \mathcal R^w(w,v)\ge\cut,
    \text{ and } bv\in\mathcal L \}.
\]
Equivalently,
\[
  \partial_a^\cut(\mathcal L)
  =
  (q_\cut^*)^{-1}
  \bigl(\partial_{[a]_\cut}(\mathcal L^\cut)\bigr)
\]
by Theorem~\ref{the:quotient-fuzzy-derivative}. Since $\mathcal L$ is regular
and $q_\cut^*$ is a homomorphism, the quotient image $\mathcal L^\cut$ is
regular. Classical derivatives preserve regularity, so
$\partial_{[a]_\cut}(\mathcal L^\cut)$ is regular over the finite quotient
alphabet $\Sigma^\cut$. Finally, inverse homomorphic images of regular languages
are regular. Hence $\partial_a^\cut(\mathcal L)$ is regular.

For a word $u = a_1 \cdots a_n$ we use the recursive definition
$\partial_u^\cut(\mathcal{L}) = \partial_{a_n}^\cut(\cdots \partial_{a_1}^\cut(\mathcal{L}) \cdots)$
and apply the previous argument inductively: at each step we take a fuzzy
derivative of a regular language and obtain a regular language again. Thus
$\partial_u^\cut(\mathcal{L})$ is regular for every $u \in \Sigma^*$ and every
cut $0 < \cut \le 1$.
\end{proof}


\noindent
In order to compute fuzzy derivatives in practice, it is convenient to identify, for a given symbol and cut value, the set of symbols that are considered similar enough to be treated as matches. We call this set the \emph{cut neighborhood} of the symbol, and define it formally as follows.


\begin{definition}\label{def:cut-neighborhood}
    For $a\in\Sigma$ and $0<\cut\le 1$, define the cut neighborhood of $a$ by
    $$
        \mathcal{S}^\cut_a = \{b \in\Sigma :  (a,b)\in\mathcal{R}_\cut\}
    $$
\end{definition}

\begin{definition}\label{def:cut-saturation}
For a language $K\subseteq\Sigma^*$, define its $\cut$-saturation by
\[
  \operatorname{Sat}_\cut(K)
  =
  \{w\in\Sigma^* \mid \exists v\in K
      \text{ such that } w\sim_\cut^{\Sigma^*}v\}.
\]
Equivalently, $\operatorname{Sat}_\cut(K)=(q_\cut^*)^{-1}(q_\cut^*(K))$.
If $R$ is a regular expression, $\operatorname{Sat}_\cut(R)$ denotes any regular
expression whose language is $\operatorname{Sat}_\cut(\mathcal L(R))$.
\end{definition}

\noindent
As in the classical case, we distinguish between the \emph{semantic} fuzzy derivative, defined directly on languages, and the \emph{syntactic} fuzzy derivative, defined on regular expressions. The syntactic formulation allows us to compute derivatives by manipulating the structure of the expression itself, without first converting it to its 
language. In the fuzzy setting, this requires taking into account all symbols in the cut neighborhood of the given symbol. The following definition formalizes the syntactic fuzzy derivative.


\begin{definition}\label{def:regex-fuzzy-derivative}
    Let $R$ be a regular expression over the alphabet $\Sigma$, let
    $a \in \Sigma$ be a symbol and let $0 < \cut \le 1$.
    The \emph{fuzzy derivative of the regular expression $R$ with respect to $a$ and $\cut$}
    is the regular expression
    $$
        \mathcal{D}^\cut_{a}(R)
    $$
    whose language satisfies
    $$
        \mathcal{L}\bigl(\mathcal{D}^\cut_a(R)\bigr)
        =
        \{v \in \Sigma^* : \exists b\in\mathcal{S}^\cut_a,\ \exists v'\in\Sigma^*
            \text{ such that } v\sim_\cut^{\Sigma^*}v'
            \text{ and } b\cdot v' \in \mathcal{L}(R)
        \}
    $$
\end{definition}

\noindent
Computationally, we can obtain this expression by first taking the head-only
fuzzy derivative and then saturating the result:
\[
  \mathcal D_a^\cut(R)
  =
  \operatorname{Sat}_\cut
  \left(
    \sum_{b\in\mathcal S_a^\cut}\mathcal D_b(R)
  \right).
\]
Here $\mathcal D_b(R)$ is the ordinary Brzozowski derivative from
Section~\ref{sec:derivatives}. Expanded structurally, the unsaturated
head-only derivative inside $\operatorname{Sat}_\cut(\cdot)$ is:
$$
\begin{array}{c}
    \mathcal{D}^{\cut,\mathrm{head}}_a(\varepsilon)=\emptyset \\[1ex]
    \mathcal{D}^{\cut,\mathrm{head}}_a(b)=
      \begin{cases}
        \varepsilon & \text{if } b \in \mathcal{S}^\cut_a,\\
        \emptyset       & \text{otherwise.}
      \end{cases}\\[1ex]
    \mathcal{D}^{\cut,\mathrm{head}}_a(\emptyset)=\emptyset \\[1ex]
    \mathcal{D}^{\cut,\mathrm{head}}_a(r\cdot s)= \mathcal{D}^{\cut,\mathrm{head}}_a(r)\cdot s + \nu(r)\cdot \mathcal{D}^{\cut,\mathrm{head}}_a(s) \\[1ex]
    \mathcal{D}^{\cut,\mathrm{head}}_a(r^*)=\mathcal{D}^{\cut,\mathrm{head}}_a(r)\cdot r^* \\[1ex]
    \mathcal{D}^{\cut,\mathrm{head}}_a(r+s)=\mathcal{D}^{\cut,\mathrm{head}}_a(r)+\mathcal{D}^{\cut,\mathrm{head}}_a(s) \\[1ex]
    \mathcal{D}^{\cut,\mathrm{head}}_a(r\:\&\:s)=\mathcal{D}^{\cut,\mathrm{head}}_a(r)\:\&\:\mathcal{D}^{\cut,\mathrm{head}}_a(s) \\[1ex]
    \mathcal{D}^{\cut,\mathrm{head}}_a(\neg r) = \neg(\mathcal{D}^{\cut,\mathrm{head}}_a(r)).
\end{array}
$$
The full fuzzy derivative is
\[
  \mathcal D_a^\cut(R)
  =
  \operatorname{Sat}_\cut\bigl(\mathcal D^{\cut,\mathrm{head}}_a(R)\bigr).
\]

\noindent
The semantic and syntactic definitions of the fuzzy derivative are intended to capture the same intuitive operation: removing a given prefix (up to similarity) from all words in the language. To confirm this alignment, we state the following correspondence theorem, which ensures that taking the fuzzy derivative at the syntactic level and then interpreting it as a language yields exactly the same result as first interpreting the expression and then applying the semantic fuzzy derivative.


\begin{theorem}[Semantic-syntactic correspondence]\label{the:two-fuzzy-derivative-mapping}
    Let $R$ be a regular expression over $\Sigma$, $u\in\Sigma^*$ and $0 < \cut \le 1$. Then
    $$
        \mathcal{L}\bigl(\mathcal{D}^\cut_u(R)\bigr)
        =
        \partial^\cut_u\bigl(\mathcal{L}(R)\bigr)
    $$
\end{theorem}

\begin{proof}[Proof.]
We prove that for all regular expressions $R$ and all words $u \in \Sigma^*$,
\[
  \mathcal{L}\!\bigl(\mathcal{D}^\cut_u(R)\bigr)
  \;=\;
  \partial^\cut_u\!\bigl(\mathcal{L}(R)\bigr).
\]
As in the crisp case, the proof has two steps: first a single symbol $a\in\Sigma$,
then an arbitrary word $u$.

\paragraph{Step 1 (single symbol).}
Fix $a \in \Sigma$. By Definition~\ref{def:regex-fuzzy-derivative} and
Definition~\ref{def:cut-saturation},
\[
  \mathcal L(\mathcal D_a^\cut(R))
  =
  \operatorname{Sat}_\cut
  \left(
    \bigcup_{b\in\mathcal S_a^\cut}
    \mathcal L(\mathcal D_b(R))
  \right).
\]
The ordinary semantic-syntactic correspondence from
Theorem~\ref{the:two-derivative-mapping} gives
$\mathcal L(\mathcal D_b(R))=\partial_b(\mathcal L(R))$. Therefore
\[
  \mathcal L(\mathcal D_a^\cut(R))
  =
  \operatorname{Sat}_\cut
  \left(
    \bigcup_{b\in\mathcal S_a^\cut}
    \partial_b(\mathcal L(R))
  \right).
\]
By Definition~\ref{fuzzyderivative}, the right-hand side is exactly
$\partial_a^\cut(\mathcal L(R))$: the union chooses the matching head symbol
$b$, and the saturation chooses all suffixes $\cut$-similar to the actual
remaining suffix. Hence
\[
  \mathcal{L}\bigl(\mathcal{D}^\cut_a(R)\bigr)
  = \partial^\cut_a\bigl(\mathcal{L}(R)\bigr)
\]
for all $a \in \Sigma$ and all regular expressions $R$.

\paragraph{Step 2 (arbitrary word).}
We prove by induction on $u \in \Sigma^*$ that
\[
  \mathcal{L}\bigl(\mathcal{D}^\cut_u(R)\bigr)
  = \partial^\cut_u\bigl(\mathcal{L}(R)\bigr).
\]
\begin{itemize}
  \item Base $u=\varepsilon$:
    $\mathcal{D}^\cut_\varepsilon(R)=R$ and
    $\partial^\cut_\varepsilon(\mathcal{L}(R))=\mathcal{L}(R)$.
  \item Step $u = va$ with $v \in \Sigma^*$ and $a \in \Sigma$:
    \[
      \begin{aligned}
      \mathcal{L}\bigl(\mathcal{D}^\cut_{va}(R)\bigr)
      &= \mathcal{L}\bigl(\mathcal{D}^\cut_a(\mathcal{D}^\cut_v(R))\bigr) \\
      &= \partial^\cut_a\bigl(\mathcal{L}(\mathcal{D}^\cut_v(R))\bigr)
           \quad\text{(by Step 1)} \\
      &= \partial^\cut_a\bigl(\partial^\cut_v(\mathcal{L}(R))\bigr)
           \quad\text{(IH on $v$)} \\
      &= \partial^\cut_{va}\bigl(\mathcal{L}(R)\bigr).
      \end{aligned}
    \]
\end{itemize}
Therefore $\mathcal{L}(\mathcal{D}^\cut_u(R))=\partial^\cut_u(\mathcal{L}(R))$
for all words $u \in \Sigma^*$, which completes the proof.
\end{proof}


\subsection*{Examples}
Assume the alphabet $\Sigma = \{a, b, c\}$ and a fuzzy similarity relation $\cR$ on $\Sigma$ defined as follows:
\[
\begin{array}{c|ccc}
\cR & a & b & c \\
\hline
a & 1 & 0.8 & 0.4 \\
b & 0.8 & 1 & 0.5 \\
c & 0.4 & 0.5 & 1 \\
\end{array}
\]
Let the cut value be $\cut = 0.7$, so the $\cut$-cut of $\cR$ includes the pairs:
\[
\cR_\cut = \{(a, a), (a, b), (b, a), (b, b), (c, c)\}
\]
We can also observe that:
\[
\begin{array}{c}
     \mathcal{S}_a^\cut=\{a, b\} \\ [1ex]
     \mathcal{S}_b^\cut=\{a, b\} \\ [1ex]
     \mathcal{S}_c^\cut=\{c\}
\end{array}
\]
\noindent
Let $\mathcal{L} = \{abc, ba, bb\}$. Then:
\begin{itemize}
    \item $\partial_a^\cut(\mathcal{L}) = \{ac, bc, a, b\}$\\
    Because:
    \begin{itemize}
        \item $(a,a) \in \cR_\cut$ and $abc \in \mathcal{L}$ leave the actual suffix $bc$; both $bc$ and $ac$ are $\cut$-similar to it.
        \item $(a,b) \in \cR_\cut$ and $ba \in \mathcal{L}$ leave the actual suffix $a$; both $a$ and $b$ are $\cut$-similar to it.
        \item $(a,b) \in \cR_\cut$ and $bb \in \mathcal{L}$ leave the actual suffix $b$; both $a$ and $b$ are $\cut$-similar to it.
    \end{itemize}

    \item $\partial_c^\cut(\mathcal{L}) = \emptyset$\\
    Because:
    \begin{itemize}
        \item Only $(c,c) \in \cR_\cut$ with $\cR(c,c) = 1 \ge \cut$
        \item But $cb, ca, cc \notin \mathcal{L}$ $\Rightarrow$ no valid $bw$ with $(c,b) \in \cR_\cut$ and $bw \in \mathcal{L}$
    \end{itemize}

    \item $\partial_b^\cut(\mathcal{L}) = \{ac, bc, a, b\}$\\
    Because:
    \begin{itemize}
        \item $(b,a) \in \cR_\cut$ and $abc \in \mathcal{L}$ leave the actual suffix $bc$; both $bc$ and $ac$ are $\cut$-similar to it.
        \item $(b,b) \in \cR_\cut$ and $ba \in \mathcal{L}$ leave the actual suffix $a$; both $a$ and $b$ are $\cut$-similar to it.
        \item $(b,b) \in \cR_\cut$ and $bb \in \mathcal{L}$ leave the actual suffix $b$; both $a$ and $b$ are $\cut$-similar to it.
    \end{itemize}
\end{itemize}

\begin{theorem}\label{the:fuzzy-derivative-set-finite}
Let $\Sigma$ be a finite alphabet and let $\mathcal{L} \subseteq \Sigma^*$ be a
regular language. For every cut value $0 < \cut \le 1$, the set of fuzzy
derivatives
\[
  \bigl\{\, \partial_u^\cut(\mathcal{L}) \;\bigm|\; u \in \Sigma^* \bigr\}
\]
is finite.
\end{theorem}

\begin{proof}
Fix $0 < \cut \le 1$ and let $\mathcal{L}$ be regular. Since $q_\cut^*$ is a
homomorphism, the quotient image $\mathcal L^\cut$ is a regular language over
the finite alphabet $\Sigma^\cut$. By Theorem~\ref{the:derivative-set-is-finite},
the set
\[
  D^\cut
  =
  \{\,\partial_x(\mathcal L^\cut) \mid x\in(\Sigma^\cut)^*\,\}
\]
is finite. By repeated use of Theorem~\ref{the:quotient-fuzzy-derivative},
every fuzzy derivative has the form
\[
  \partial_u^\cut(\mathcal L)
  =
  (q_\cut^*)^{-1}
  \bigl(\partial_{q_\cut^*(u)}(\mathcal L^\cut)\bigr).
\]
Thus the possible fuzzy derivatives are inverse images of elements of the finite
set $D^\cut$. Hence
$\{\partial_u^\cut(\mathcal{L}) \mid u \in \Sigma^*\}$ is finite.
\end{proof}



\begin{theorem}\label{the:regex-fuzzy-derivative-types-finite}
Let $\Sigma$ be a finite alphabet and let $R$ be a regular expression over
$\Sigma$. For every cut value $0 < \cut \le 1$, the set of derivative languages
\[
  \bigl\{\, \mathcal{L}(\mathcal{D}^\cut_{w}(R)) \;\bigm|\; w \in \Sigma^* \bigr\}
\]
is finite.
\end{theorem}

\begin{proof}
Let $0 < \cut \le 1$ be fixed and $\mathcal{L} = \mathcal{L}(R)$.
By Theorem~\ref{the:two-fuzzy-derivative-mapping}, for every word $w \in \Sigma^*$,
\[
  \mathcal{L}\bigl(\mathcal{D}^\cut_w(R)\bigr)
  \;=\;
  \partial^\cut_w\bigl(\mathcal{L}(R)\bigr)
  \;=\;
  \partial^\cut_w(\mathcal{L}).
\]
Hence
\[
  \bigl\{\, \mathcal{L}(\mathcal{D}^\cut_{w}(R)) \mid w \in \Sigma^* \bigr\}
  \;=\;
  \bigl\{\, \partial^\cut_w(\mathcal{L}) \mid w \in \Sigma^* \bigr\},
\]
and the right-hand set is finite by
Theorem~\ref{the:fuzzy-derivative-set-finite}. This proves the claim.
\end{proof}



\begin{theorem}\label{the:fuzzy-dfa-correctness}
Let $r$ be a regular expression over $\Sigma$, let $0 < \cut \le 1$, and let
$M^\cut = \langle Q^\cut, \Sigma, \delta^\cut, q_0^\cut, F^\cut \rangle$
be the DFA constructed from $r$ using fuzzy derivatives (for the fixed cut
value $\cut$).
Write
\[
  \mathcal L^\cut_\Sigma(r)
  =
  (q_\cut^*)^{-1}\bigl(q_\cut^*(\mathcal L(r))\bigr)
  =
  \operatorname{Sat}_\cut(\mathcal L(r))
\]
for the concrete language of words that are $\cut$-similar to some word denoted
by $r$. Then $M^\cut$ and $r$ define the same concrete $\cut$-language:
\[
  \mathcal{L}(M^\cut) \;=\; \mathcal{L}^\cut_\Sigma(r).
\]
Equivalently, for every word $w \in \Sigma^*$,
\[
  M^\cut \text{ accepts } w
  \;\;\Longleftrightarrow\;\;
  w \in \mathcal{L}^\cut_\Sigma(r).
\]
\end{theorem}

\begin{proof}[Proof.]
Let $\delta^{\cut *} : Q^\cut \times \Sigma^* \to Q^\cut$ be the standard
extension of $\delta^\cut$ to words:
\[
  \delta^{\cut *}(q,\varepsilon) = q,
  \qquad
  \delta^{\cut *}(q,wa) = \delta^\cut\bigl(\delta^{\cut *}(q,w),a\bigr)
  \text{ for } w \in \Sigma^*, a \in \Sigma.
\]

We first show by induction on $w \in \Sigma^*$ that
\[
  \delta^{\cut *}(q_0^\cut,w) = \bigl[\,\mathcal{D}^\cut_w(r)\,\bigr].
\]

\emph{Base case:} For $w = \varepsilon$ we have
$q_0^\cut = [r] = [\mathcal{D}^\cut_\varepsilon(r)]$, so the claim holds.

\emph{Inductive step:} Let $w = ua$ with $u \in \Sigma^*$ and $a \in \Sigma$.
By the induction hypothesis,
$\delta^{\cut *}(q_0^\cut,u) = [\mathcal{D}^\cut_u(r)]$.
Then, by the definition of $\delta^\cut$,
\[
  \delta^{\cut *}(q_0^\cut,ua)
  = \delta^\cut\bigl(\delta^{\cut *}(q_0^\cut,u),a\bigr)
  = \delta^\cut\bigl([\mathcal{D}^\cut_u(r)],a\bigr)
  = \bigl[\,\mathcal{D}^\cut_a(\mathcal{D}^\cut_u(r))\,\bigr]
  = \bigl[\,\mathcal{D}^\cut_{ua}(r)\,\bigr],
\]
using the rule $\mathcal{D}^\cut_{ua}(r) = \mathcal{D}^\cut_a(\mathcal{D}^\cut_u(r))$.
This proves the claim.

Now let $w \in \Sigma^*$. Then:
\[
\begin{aligned}
w \text{ is accepted by } M^\cut
&\Longleftrightarrow \delta^{\cut *}(q_0^\cut,w) \in F^\cut \\[0.5ex]
&\Longleftrightarrow \delta^{\cut *}(q_0^\cut,w) = [\mathcal{D}^\cut_w(r)]
             \text{ and } [\mathcal{D}^\cut_w(r)] \in F^\cut
             \quad\text{(by the claim above and def.\ of $F^\cut$)} \\[0.5ex]
&\Longleftrightarrow \nu\bigl(\mathcal{D}^\cut_w(r)\bigr) = \varepsilon \\[0.5ex]
&\Longleftrightarrow \varepsilon \in \mathcal{L}\bigl(\mathcal{D}^\cut_w(r)\bigr) \\[0.5ex]
&\Longleftrightarrow \varepsilon \in \partial^\cut_w\bigl(\mathcal{L}(r)\bigr)
     \quad\text{(by Theorem~\ref{the:two-fuzzy-derivative-mapping})} \\[0.5ex]
&\Longleftrightarrow w \in \mathcal{L}^\cut_\Sigma(r)
     \quad\text{(by the definition of $\mathcal{L}^\cut_\Sigma(r)$).}
\end{aligned}
\]
Hence $\mathcal{L}(M^\cut) = \mathcal{L}^\cut_\Sigma(r)$, as required.
\end{proof}
